% Budgeted ISAC manuscript insert generated from the 2026-07-09 evidence package.
% This file is intended as a controlled insert for main.tex after the paper
% narrative is updated from the earlier Phase10 MARL evidence line.

\subsection{Density-Adaptive Budgeted ISAC Access}
\label{subsec:budgeted_isac}

The preceding ISAC candidate mechanism substantially reduces empty-beam probing, but it can also concentrate many UAVs on a small set of high-confidence beam cells.
In dense single-hop swarms this creates a second bottleneck: after empty beams are suppressed, simultaneous transmissions toward popular candidate sectors generate receiver-side collisions.
We therefore introduce a density-adaptive budgeted ISAC access rule that keeps the ISAC candidate executor but constrains the transmit probability using only local memory and configuration parameters.
The rule is not an oracle; it uses no undiscovered-neighbor position, identity, or global topology.
It is also useful as a rule expert for the later learned access gate.

Let $\mathcal{C}_i^t$ be the ISAC candidate set of UAV $i$ in slot $t$, $d_i^t$ its discovered degree, and $d_{\rm tar}$ the target local degree.
Define the normalized degree deficit
\begin{equation}
  \Delta_i^t = \frac{\max\{0,d_{\rm tar}-d_i^t\}}{\max\{1,d_{\rm tar}\}}.
\end{equation}
The local candidate pressure is
\begin{equation}
  \rho_{i,{\rm cand}}^t =
  \min\left\{1,\frac{|\mathcal{C}_i^t|}{\max\{1,d_{\rm tar}\}}\right\}.
\end{equation}
For candidate beams $b\in\mathcal{C}_i^t$, node $i$ maintains belief $q_i^t(b)$, success count $s_i^t(b)$, failure count $f_i^t(b)$, and collision-failure count $c_i^t(b)$.
The candidate evidence, failure pressure, and collision pressure are
\begin{align}
  e_i^t &=
  \frac{1}{|\mathcal{C}_i^t|}
  \sum_{b\in\mathcal{C}_i^t}
  \left[q_i^t(b)+0.15\log(1+s_i^t(b))\right]_{0}^{1},\\
  \rho_{i,{\rm fail}}^t &=
  \frac{1}{|\mathcal{C}_i^t|}
  \sum_{b\in\mathcal{C}_i^t}
  \frac{f_i^t(b)}{\max\{1,s_i^t(b)+f_i^t(b)\}},\\
  \rho_{i,{\rm col}}^t &=
  \frac{1}{|\mathcal{C}_i^t|}
  \sum_{b\in\mathcal{C}_i^t}
  \frac{c_i^t(b)}{\max\{1,s_i^t(b)+c_i^t(b)\}},
\end{align}
where $[\cdot]_0^1$ denotes clipping to $[0,1]$.
If the candidate set is empty, these pressures are set to zero.
Let the codebook density factor be
\begin{equation}
  \chi = \sqrt{1 + 4 \frac{N}{|\mathcal{B}|} d_{\rm tar}},
\end{equation}
where $N$ is the swarm size and $|\mathcal{B}|$ is the number of three-dimensional beam cells.
The budgeted transmit probability is
\begin{equation}
\begin{split}
  p_{i,{\rm tx}}^t =
  \Big[
  \frac{0.48}{\chi}
  +0.08\Delta_i^t
  +0.05 e_i^t
  -0.08\rho_{i,{\rm cand}}^t
  -0.18\rho_{i,{\rm col}}^t
  -0.08\rho_{i,{\rm fail}}^t
  -0.08\mathbb{1}\{\Delta_i^t=0\}
  \Big]_{0.16}^{0.45}.
\end{split}
\label{eq:budgeted_tx_probability}
\end{equation}
The idle and receive probabilities are
\begin{align}
  p_{i,{\rm idle}}^t &=
  [0.01+0.02(1-\Delta_i^t)+0.04\rho_{i,{\rm col}}^t]_{0.01}^{0.09},\\
  p_{i,{\rm rx}}^t &= 1-p_{i,{\rm tx}}^t-p_{i,{\rm idle}}^t .
\end{align}
This rule lowers transmit probability when the swarm is dense relative to the beam codebook, when local candidates are crowded, or when recent candidate beams have accumulated collision evidence.
It raises activity only when the node has a local degree deficit and credible candidate evidence.

\paragraph*{Result summary}
In the N=100, B=10 long-horizon transfer setting, the budgeted rule achieves discovery rate 0.712, collision count 1347.0, and collision-penalized discovery 0.560.
The corresponding Wang-style ISAC baseline achieves 0.681, 1343.3, and 0.536, while the non-budgeted collision-aware ISAC rule achieves 0.767, 4676.7, and 0.395.
Thus the budgeted rule preserves most of the raw-discovery gain of aggressive ISAC access while recovering the collision efficiency of conservative Wang-style access.
At N=100, B=15, it achieves collision-penalized discovery 0.245 versus 0.227 for Wang-style ISAC and 0.116 for non-budgeted collision-aware ISAC.

\paragraph*{Implication for MARL}
The current learned BC-MARL policy improves raw discovery relative to earlier MARL runs but produces excessive collisions under N=100 transfer.
The budgeted ISAC rule should therefore be treated as the access-control expert to distill into a learned decentralized gate, preferably with a collision-budget or Lagrangian penalty.
The learning target is not to rediscover beam candidates from scratch, but to learn when to throttle or activate narrow-beam access after ISAC has removed most empty sectors.
